### 2023-11-29 thesis/MSECE_thesis.tex
Final version of thesis/refs.

\section*{Abstract}\label{sec:abstract}

\section{Introduction}\label{sec:introduction}
There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Automatic speech recognition and
language modeling technologies, in particular, have made tremendous advances over the past two decades achieving performance levels good enough to
be brought to consumer facing products such as personal assistants, search engines, phones, televisions, etc. Due to the success in the general
domain, applications have begun to spin-off into problem-specific domains, such as (and most notably for this work) aviation. For example, several
speech recognition systems with rule-based natural language inference and semantic extraction algorithms have been implemented at EUROCONTROL
simulation centers in Europe to train air traffic controllers and study the potential effects of AI augmented air traffic management (ATM) systems
on controller workloads. Preliminary studies have found that AI augmented ATM systems not only reduce controller workloads but also reduce fuel
consumption of aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times thereby reducing fuel
consumption~\cite{helmke_quantifying_2017}.

At the time of this writing, the current focus of general machine learning algorithms in the automatic speech recognition (ASR) and language modeling
domains has been on a variety of deep learning methods/architectures, namely; recurrent, convolutional, and transformer neural networks due to the
availability of large corpora such as LibriSpeech~\cite{panayotov_librispeech_2015} and Mozilla Common Voice~\cite{ardila_common_2020} for ASR and
WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011} corpora for language modeling. In contrast to the general domain, the aviation
domain still prefers, and sees much success with, traditional machine learning models for both ASR and language modeling applications. Typically, the
most prevalent of the traditional models are Hidden Markov models and Gaussian Mixture models, in conjunction with word lattices, N-grams, and/or
rule-based approaches depending on the application and operating environment of the models. The most likely reason for this divergence between the
domains is the difference and relative lack of labeled data in the aviation domain. By comparison, one of the largest corpora in the general domain is
LibriSpeech with approximately 1000 hours of labeled data whereas, for aviation, the largest corpus is Air Traffic Control Complete with approximately
26 hours of labeled data\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are
    labeled.}. This disparity in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to
specialized pronunciation for some words to enhance understandability in a radiotelephonic (R/T) medium~\cite{paltridge_handbook_2013}, making it
significantly more difficult to transcribe/label data as it requires transcribers to have domain knowledge and experience in order to create effective
transcriptions. The medium of communication along with the environment in which the communication occurs i.e.~a noisy and possibly low fidelity
medium\footnote{Depending on the equipment used to transmit and receive R/T communications.} combined with a noisy and active communication
environment---an active flight deck (influenced by wind, engine noise, etc.~) or an active Air Traffic Control (ATC) tower (other people/controllers
conversing or communicating in the background)---and a rapid rate of speech~\cite{paltridge_handbook_2013}, further contributing to the difficulty in
understanding and therefore transcribing aviation English.

@@

\section{Literature Review}\label{sec:lit_review}
% from thesis proposal; revise and add to this as more references are used
% transformer history (background)
The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved
state-of-the-art (28.4 BLEU on WMT 2014 English-to-German; 41.8 BLEU on WMT English-to-French datasets)\cite{vaswani_attention_2017}. Transformer
architectures have been found to be extremely effective at learning representations and understandings of languages to predict token probabilities as
opposed to transforming one language into another\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at
other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks,
respectively)\cite{lewis_bart_2019,radford_improving_2018}.

% ASR context (suggest some need for semi-supervised approaches and therefore LMs)
End-to-end generative models for automatic speech recognition models have made significant progress in recent years with current state-of-the-art
models achieving WERs as low as 2\% on LibriSpeech test sets\cite{han_contextnet_2020,kriman_quartznet_2020,baevski_wav2vec_2020,li_jasper_2019}.
This has led to the development of ASR models for the aviation domain, specifically, in air traffic control communications\cite{badrinath_automatic_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,srinivasamurthy_semi-supervised_2017}. However, due to the lack of
transcribed data in the aviation domain\cite{zuluaga-gomez_automatic_2020,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022,smidl_air_2019}
ASR models maintain relatively high WERs compared to their counterparts in the more generalized ASR domain\cite{zuluaga-gomez_automatic_2020,badrinath_automatic_2022}. Transfer learning has even yielded limited results in this domain (depending on model
architecture and dataset quality)\cite{badrinath_automatic_2022,zuluaga-gomez_automatic_2020}.

% Unsupervised/semi-supervised in general and LMs commonly used with them
Unsupervised and semi-supervised methodologies have become popular recently in attempts to address limited data availability
and develop new approaches towards modeling human speech (notably, wav2vec has achieved state-of-the-art performance with
very little training data)\cite{baevski_wav2vec_2020,badrinath_automatic_2022,srinivasamurthy_semi-supervised_2017,zuluaga-gomez_contextual_2021}.
Language models are an integral part semi-supervised learning. They are used to obtain a certainty score (or uncertainty
score, as the case may be) for the predicted text, these are usually either word lattices or N-gram models\cite{badrinath_automatic_2022,srinivasamurthy_semi-supervised_2017,zuluaga-gomez_contextual_2021}.
While these have proven to be adequate for most self-supervised learning tasks, they are hardly state-of-the-art.
Wav2vec, the current state-of-the-art unsupervised model (and possibly top performing overall), uses a contrastive process
between convolutional feature extraction and transformer context representations\cite{baevski_wav2vec_2020}.

% LMs and NLP in aero. domain
Various natural language processing methods have been applied to the aviation domain to help deal with miscommunications
and try to mitigate safety incidents\cite{ragnarsdottir_language_2003,tanguy_natural_2016,madeira_machine_2021}.
Some machine learning approaches have been implemented to analyze the text in aviation incident and safety reports to predict
contributing factors and topic models\cite{tanguy_natural_2016,madeira_machine_2021}. An ASR system with NLP post-processing
has also been proposed to analyze Air-Traffic communications and condense significant features (e.g.~weather, runway, and
heading info) into an XML language structure\cite{ragnarsdottir_language_2003}. Transformer language models such as BERT,
RoBERTa, and DeBERTa have been applied to notice to airmen (NOTAM) messages to perform named entity recognition (NER) tasks,
translation tasks (between notations; e.g.~NOTAM notation to Airlang to make parsing tasks easier) and reduce the workload
for pilots during long-haul flights\cite{arnold_knowledge_2022}. Transformer models have also been used for speaker role
identification tasks in the aviation domain (e.g.~identifying pilot versus controller in communications), specifically,
a pretrained BERT model was used and fine-tuned on problem specific data and compared to other models that performed
well at speaker and role identification tasks in general\cite{guo_comparative_2022}.

@@

All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used
    although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts parsing from the
corpora-specific formats where necessary i.e.~Lisp in ATCC, XML in ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations
are removed where necessary\footnote{A copy of the original data is made, then the annotations are removed from the copy.}, and the resulting text
is written to a file in the standard corpus format (an ASCII text file with one line in the file corresponding to one sample from the corpus) to a
file.

@@

@@

The pretraining methodology introduced by RoBERTa eliminated the NSP pretraining task that was originally used in BERT and opted instead to increase
the quantity of training data along with the batch sizes in addition to training the model longer. Lastly, there is a small difference in the
construction of input sequences as compared to BERT; as with BERT, input sequences always begin with a classification token (\lstinline|[CLS]|) and
individual segments are separated by a separation token (\lstinline|[SEP]|). Rather than having sequences end with a separation token as well, they
are terminated with an end of sequence token (\lstinline|[EOS]|). See equation \ref{eq:roberta_input_example} modified from Liu et
al.~\cite{liu_roberta_2019}, below.

@@

As mentioned previously, perplexity and cross-entropy loss (often shortened to loss) are used to track the performance of the model on the training
and validation sets, per epoch, while training. Cross-entropy loss is defined by the following equation:

@@

With the metrics above defined, an analysis of the results of the model pretraining can be performed. The objective of the pretraining procedure is to
build the models' understanding of the structure of the text, in this case transcripts of air-traffic communications. By tracking the average
cross-entropy loss (loss) and perplexity (PPL) during training, we can estimate how well (and how correctly) the models are predicting tokens and if
the models are effectively ``understanding'' token placements in sequences.

>>>>

\section{Abstract}\label{sec:abstract}

\section{Introduction}\label{sec:introduction}
There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Natural language processing
and language modeling technologies, in particular, have made tremendous advances over the past two decades achieving performance levels good enough to
be brought to consumer facing products such as personal assistants, search engines, phones, televisions, etc. Due to the success in the general
domain, applications have begun to spin-off into problem-specific domains, such as (and most notably for this work) aviation. For example, several
speech recognition systems with rule- and machine learning-based natural language processing, inference, and semantic extraction algorithms have been
implemented at EUROCONTROL simulation centers in Europe to train air traffic controllers and study the potential effects of AI augmented air traffic
management (ATM) systems on controller workloads. Preliminary studies have found that AI augmented ATM systems not only reduce controller workloads
but also reduce fuel consumption of aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times thereby reducing fuel
consumption~\cite{helmke_quantifying_2017}.

At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a
variety of deep learning methods/architectures, namely; recurrent, convolutional, and transformer neural networks due to the availability of large
natural language corpora corpora such as WikiText~\cite{merity_pointer_2016} and IMDB~\cite{maas_learning_2011}. In contrast to the general domain,
the aviation domain still prefers, and sees much success with, traditional machine learning models for language modeling applications. Typically, the
most prevalent of the traditional models are Hidden Markov models, Gaussian Mixture models, and/or in conjunction with word lattices, N-grams, and/or
rule-based approaches depending on the application and operating environment of the models. The most likely reason for this divergence between the
domains is the difference and relative lack of labeled data in the aviation domain. By comparison, the largest corpora in the general domain is either
WikiText or the BookCorpus which contain trillions of tokens whereas, for aviation, the largest corpus is Air Traffic Control Complete with
approximately 26 hours of labeled data amounting to just over 300,000 tokens\footnote{Air Traffic Control Complete has approximately 70 hours of
    audio data, however, about 26 hours of the 70 are labeled.}. This disparity in data availability is exacerbated by the fact that aviation English
uses a specialized vocabulary in addition to specialized pronunciation for some words to enhance understandability in a radiotelephonic (R/T)
medium~\cite{paltridge_handbook_2013}, making it significantly more difficult to transcribe/label data as it requires transcribers to have domain
knowledge and experience in order to create effective transcriptions. The medium of communication along with the environment in which the
communication occurs i.e.~a noisy and possibly low fidelity medium\footnote{Depending on the equipment used to transmit and receive R/T
    communications.} combined with a noisy and active communication environment---an active flight deck (influenced by wind, engine noise, etc.~) or
an active Air Traffic Control (ATC) tower (other people/controllers conversing or communicating in the background)---and a rapid rate of speech
\cite{paltridge_handbook_2013}, further contributing to the difficulty in understanding and therefore transcribing aviation English.

@@

\section{Literature Review}\label{sec:lit_review}
% from thesis proposal; revise and add to this as more references are used
% transformer history (background)
The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved
state-of-the-art (28.4 BLEU on WMT 2014 English-to-German; 41.8 BLEU on WMT English-to-French datasets)\cite{vaswani_attention_2017}. Transformer
architectures have been found to be extremely effective at learning representations and understandings of languages to predict token probabilities as
opposed to transforming one language into another\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at
other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks,
respectively)\cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~proposed pretraining methods such as masked language
modeling and next sentence prediction to improve natural language understanding for downstream tasks \cite{devlin_bert_2019}, which was further
expanded upon by Liu et al.~by training longer, with larger batch sizes, and removing the next sentence prediction task \cite{liu_roberta_2019}.

% tokenizer algorithms
In language modeling and natural language processing tasks more generally, the out-of-vocabulary (OOV) problem is prevalent and has been addressed by
the introduction of tokenizer algorithms. WordPiece and Byte Pair Encoding have been used to solve OOV problems while simultaneously minimizing
vocabulary sizes and maximizing the likelihood of tokens in sequences \cite{wu_googles_2016,schuster_japanese_2012,sennrich_neural_2016}.

% LMs and NLP in aero. domain
Various natural language processing methods have been applied to the aviation domain to help deal with miscommunications
and try to mitigate safety incidents\cite{ragnarsdottir_language_2003,tanguy_natural_2016,madeira_machine_2021}.
Some machine learning approaches have been implemented to analyze the text in aviation incident and safety reports to predict
contributing factors and topic models\cite{tanguy_natural_2016,madeira_machine_2021}. An ASR system with NLP post-processing
has also been proposed to analyze Air-Traffic communications and condense significant features (e.g.~weather, runway, and
heading info) into an XML language structure\cite{ragnarsdottir_language_2003}. Transformer language models such as BERT and RoBERTa have been applied
to notice to airmen (NOTAM) messages to perform named entity recognition (NER) tasks, translation tasks (between notations; e.g.~NOTAM notation to
Airlang to make parsing tasks easier) and reduce the workload for pilots during long-haul flights\cite{arnold_knowledge_2022}. Transformer models
have also been used for speaker role identification tasks in the aviation domain (e.g.~identifying pilot versus controller in communications),
specifically, a pretrained BERT model was used and fine-tuned on problem specific data and compared to other models that performed well at speaker
and role identification tasks in general\cite{guo_comparative_2022}.

% ASR context (suggest some need for semi-supervised approaches and therefore LMs)
Automatic speech recognition is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification,
etc. A majority of these approaches use machine learning approaches (as opposed to deep learning) where acoustic, pronunciation, and language
modeling is used to transcribe speech
\cite{guo_comparative_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,badrinath_automatic_2022,hofbauer_atcosim_2008,helmke_quantifying_2017}.
Language models have been used in the general domain (both neural and machine learning-based approaches) to boost the performance of speech
recognition models \cite{han_contextnet_2020,kriman_quartznet_2020,majumdar_citrinet_2021}. Machine learning-based language models, usually N-grams
and word lattices have also been used in the aviation domain alongside semi-supervised approaches to automatic speech recognition to leverage
unlabeled data, however, none of the developed methodologies use neural language models for their applications
\cite{zuluaga-gomez_contextual_2021,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022}.

% differences in language (transcripts vs literary english), narrow model selection
The use of the English language in aviation has been shown to be sufficiently different enough to be classified under its own category (aviation
English) \cite{paltridge_handbook_2013}. The issue with using pretrained checkpoints of transformer-based language models is characterizing the
similarity (or, conversely, the difference) between literary English and aviation English. If the two domains are sufficiently different, then
training the pretrained models on aviation English data would amount to transfer learning between specific domains. Transfer learning between text
domains has been shown to be possible even with data volumes significantly lower than the original data used to train the models
\cite{raffel_exploring_2020}. Although, Yadlowsky et al.~recently showed that if problem domains are sufficiently different, model pretraining on data
considered out-of-domain will have negative effects on model performance \cite{yadlowsky_pretraining_2023}.

@@

All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used
    although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts by parsing from the
corpora-specific formats where necessary i.e.~Lisp in ATCC, XML in ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations
are removed where necessary\footnote{A copy of the original data is made, then the annotations are removed from the copy.}, and the resulting text
is written to a file in the standard corpus format (an ASCII text file with one line in the file corresponding to one sample from the corpus) to a
file.

@@

One of the intended downstream use cases of the language models developed from this work is decoding speech recognition predictions into transcripts,
so including as much data as possible regarding the realization of spoken words is useful given the context. The \lstinline|[HES]| token is used as a
catch-all, fallback token in the case that no knowledge can be extracted about the utterance from the transcript.

@@

The pretraining methodology introduced by RoBERTa eliminated the NSP pretraining task that was originally used in BERT and opted instead to increase
the quantity of training data along with the batch sizes in addition to training the model longer. Lastly, there is a small difference in the
construction of input sequences as compared to BERT; input sequences always begin with a classification token (\lstinline|[CLS]|) and individual
segments are separated by a separation token (\lstinline|[SEP]|). Rather than having sequences end with a separation token as well, they are
terminated with an end of sequence token (\lstinline|[EOS]|). See equation \ref{eq:roberta_input_example} modified from Liu et
al.~\cite{liu_roberta_2019}, below.

@@

\subsection{Performance Metrics}\label{sec:performance_metrics}
As mentioned previously, perplexity and cross-entropy loss (often shortened to loss) are used to track the performance of the model on the training
and validation sets, per epoch, while training. Cross-entropy loss is defined by the following equation:

@@

\subsection{Pretraining Results}\label{sec:pretraining_results}
With the metrics above defined, an analysis of the results of the model pretraining can be performed. The objective of the pretraining procedure is to
build the models' understanding of the structure of the text, in this case transcripts of air-traffic communications. By tracking the average
cross-entropy loss (loss) and perplexity (PPL) during training, we can estimate how well (and how correctly) the models are predicting tokens and if
the models are effectively ``understanding'' token placements in sequences.
